---
title: "Hearing distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hearing distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`sonify_data()` plays rows in order, which works when the order means
something: games in a season, sleds crossing a finish line. But a lot of
data isn't a sequence. For the ages of houses in a state, the order of the
rows is arbitrary. What matters is the **distribution**: how many houses
there are of each age.

A histogram shows a distribution by sorting values into bins and drawing a
bar for each one. `sonify_histogram()` does the same thing for your ears. It
sweeps across the bins from lowest to highest, playing one note per bin.
The more rows in a bin, the higher and louder the note. Empty bins are
silent.

```{r setup, message = FALSE}
library(soundeR)
library(dplyr)
```

## Houses in Nebraska

`ne_house_years` counts the single-family houses in Nebraska by the year they
were built, from parcel records. (About 13,000 parcels have no build year,
so they aren't here.)

```{r}
ne_house_years
```

The data is already counted, one row per year, so the counts go to
`weight`. (With raw data, one row per house, you'd leave `weight` out.)
Let's start with one bin per decade:

```{r}
ne_house_years |>
  sonify_histogram(year_built, weight = houses, binwidth = 10, instrument = "marimba")
```

The whole sweep takes 20 seconds, one note per decade. The 1800s start low
and climb to a boom in the first decades of the 1900s. Then comes a sharp
drop in the 1930s, the Depression, followed by a long rise after World War
II that peaks in the 1970s, a dip in the 1980s, another peak in the 2000s,
and a fall after the 2008 housing crash.

`notes()` shows each bin's edges and count:

```{r}
ne_house_years |>
  sonify_histogram(year_built, weight = houses, binwidth = 10) |>
  notes() |>
  select(bin_start, bin_end, note, value)
```

Each bin includes its start but not its end, so the 1900 bin covers 1900
through 1909.

## The bins change what you hear

Here's the same data with one bin per year:

```{r}
ne_house_years |>
  sonify_histogram(year_built, weight = houses, binwidth = 1, instrument = "marimba")
```

It sounds completely different. Through the late 1800s and early 1900s,
jarring high notes jump out every five and ten years. Those are years like
1900, which has 19,205 houses, compared with 730 in 1901:

```{r}
ne_house_years |>
  filter(year_built >= 1898, year_built <= 1906)
```

Did Nebraska really build 26 times as many houses in 1900 as in 1901? Before
1950, 42% of the houses in this data have a year ending in 0, where you'd
expect about 10%. From 1980 on, the round-number spikes mostly disappear. A
pattern that regular says something about how the years got into the
records, not just about when houses went up. Why it happened isn't
documented yet. It's the kind of question to take to the people who keep
the data before you report a number like "19,205 houses built in 1900."

Neither version is wrong. Decade bins show the overall shape and hide the
rounding. Year bins show the rounding and bury the shape. Picking a bin
width is a choice, in sound just as in a chart, so try more than one.

## A few big bins can drown out the rest

When a handful of bins are far bigger than the others, most notes crowd
together at the bottom of the range. `log = TRUE` bases pitch on the
logarithm of the counts, which spreads out the small ones:

```{r}
ne_house_years |>
  sonify_histogram(year_built, weight = houses, binwidth = 1, log = TRUE, instrument = "marimba")
```

Now you can hear the handful of houses from the early 1800s rising toward
the 1900s instead of sitting on one low note. The trade-off is that
differences between big bins sound smaller than they are.

## A smooth curve

A histogram's bins are all-or-nothing: a house built in 1909 counts toward
one bar and a house built in 1910 toward the next. A density curve, like
`ggplot2::geom_density()`, smooths the distribution instead. `sonify_density()`
plays that curve as one continuous tone that glides up where there are more
houses and down where there are fewer:

```{r}
ne_house_years |>
  sonify_density(year_built, weight = houses, adjust = 4)
```

`adjust` sets how smooth the curve is, the same as in `geom_density()`.
With `adjust = 4` you hear three rises: to a peak around 1916, a dip in the
late 1930s, a long climb to about 1971, a sag around 1992, and a last peak
around 2001. With the default of `adjust = 1`, the curve is detailed enough
to wobble at 1901, 1910 and 1920, the round-number years again:

```{r}
ne_house_years |>
  sonify_density(year_built, weight = houses)
```

The glide uses soundeR's built-in tones. To hear the curve on a real
instrument, add `glide = FALSE`, which plays it as a quick run of notes:

```{r}
ne_house_years |>
  sonify_density(year_built, weight = houses, adjust = 4, glide = FALSE, instrument = "marimba")
```

## Watch the edges of your data

The last two years in this data, 2025 and 2026, were only partly recorded
when it was compiled: 408 houses and 1. Played as they are, they sound like
the housing market fell off a cliff. Leaving out the incomplete years is
more honest:

```{r}
ne_house_years |>
  filter(year_built < 2025) |>
  sonify_histogram(year_built, weight = houses, binwidth = 5, instrument = "marimba")
```

## Raw data works too

Without `weight`, every row counts once. Here's how the Husker basketball
margins were spread across the 2025-26 season, in five-point bins:

```{r}
husker_games |>
  sonify_histogram(point_margin, binwidth = 5, instrument = "xylophone", length = 10)
```

The sweep starts with the seven losses. They're spread thin across the left
side, with a silent gap where no game landed, so their notes are low:
remember, pitch here is how many games fell in a bin, not how big the margin
was. The notes climb for the most common results, wins by 0 to 20 points,
then scatter through the blowouts, including five wins by 30 or more.

## Seeing it

`sonify_video()` draws a sonified histogram as a histogram, with bars that
fill in as the sweep passes them. A density curve fills in the same way.

```{r, eval = FALSE}
ne_house_years |>
  sonify_histogram(year_built, weight = houses, binwidth = 1, instrument = "marimba") |>
  sonify_video(
    "house-years.mp4",
    title = "Houses built in Nebraska, by year",
    subtitle = "One bar per year. Listen for the round numbers.",
    x_label = "Year built",
    y_label = "Houses"
  )
```

## Things to try

* Try `binwidth = 5`. Do the round-number spikes go away, or just move?
* Which do you trust more to describe housing in Nebraska, the year version
  or the decade version? What would you tell a reader about each?
* Leave out the years before 1950 with `filter()`. How does the rest of the
  distribution sound without the round-number years?
* Find a column in your own data where row order doesn't matter, and listen
  to its distribution.
