---
title: "Your first sonification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your first sonification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A chart turns numbers into shapes you can see. A **sonification** turns
numbers into sounds you can hear. Higher numbers can become higher notes,
the gap between two events can become the silence between two notes, and a
whole season of games can become a short tune.

soundeR does this with one function, `sonify_data()`. It works like the
rest of the tidyverse: give it a data frame, tell it which columns to
listen to, and print the result. In RStudio or Positron an audio player
appears in the Viewer pane. In a Quarto or R Markdown document, like this
one, the player appears on the page.

```{r setup, message = FALSE}
library(soundeR)
library(dplyr)
```

## Real instruments (one-time setup)

soundeR can play 128 real instruments, from piano to xylophone to cello.
They need a one-time download of instrument sounds, about 30 MB:

```{r, eval = FALSE}
sound_setup()
```

If you skip this, everything still works. soundeR uses its own built-in
sounds instead and tells you when it does.

## Fractions of a second

In 2010, during the Vancouver Winter Olympics, The New York Times published
"Fractions of a Second: An Olympic Musical." For each event, it played one
note for every athlete at the moment they crossed the finish line, spaced
by how far they finished behind the winner. You could *hear* how close the
races were.

We can rebuild that with `luge_finals`, the results of every luge event at
the 2026 Winter Olympics in Milan and Cortina d'Ampezzo:

```{r}
luge_finals |>
  select(event, rank, athlete, country, total_time, behind)
```

`behind` is how many seconds each sled finished behind the winner of its
event. Let's listen to the men's singles final:

```{r}
mens_singles <- luge_finals |>
  filter(event == "Men's singles")

mens_singles |>
  sonify_data(time = behind, time_scale = 1, instrument = "piano")
```

There's no pitch here. Every note is the same. Only two things matter:

* `time = behind` decides **when** each note plays. The winner, 0 seconds
  behind, plays first.
* `time_scale = 1` plays it in **real time**: one second of data is one
  second of sound.

Max Langenhan won by 0.596 seconds, and you can hear the gap before the
second note. Now try the men's doubles, where gold and silver were
separated by 0.068 seconds. In real time those first notes nearly blur
together, so slow it down four times with `time_scale = 4`:

```{r}
luge_finals |>
  filter(event == "Men's doubles") |>
  sonify_data(time = behind, time_scale = 4, instrument = "piano")
```

To hear every event, one after another like the original, use
`sequence`. It plays each group in turn, with a second of silence between
them:

```{r}
luge_finals |>
  sonify_data(time = behind, time_scale = 1, sequence = event, instrument = "piano")
```

### Working it out yourself

`luge_finals` already has a `behind` column, but real data usually won't.
You can compute it with `group_by()` and `mutate()`:

```{r}
luge_finals |>
  group_by(event) |>
  mutate(seconds_behind = total_time - min(total_time)) |>
  sonify_data(time = seconds_behind, time_scale = 1, sequence = event, instrument = "piano")
```

Grouping your data doesn't change the sound by itself. `sequence = event`
is what plays the events one at a time. If you pass grouped data without
`sequence`, soundeR reminds you of that.

## What did the data become?

Every row became a note. `notes()` shows you exactly which one, the same way
you'd look at the data behind a chart:

```{r}
race <- mens_singles |>
  sonify_data(time = behind, time_scale = 1, instrument = "piano")

notes(race)
```

`row` matches the row of your data, `onset` is the second the note starts,
`note` is its name (C5 is the C an octave above middle C), and `value` is
the number it came from.

## A season you can hear

`husker_games` holds every game of Nebraska men's basketball's 2025-26
season, the best in program history: 28 wins, a 20-0 start, and the
program's first NCAA tournament wins.

```{r}
husker_games |>
  select(game_number, date, opponent, husker_score, opponent_score, point_margin)
```

This time, let's map `point_margin` to **pitch**. Bigger wins are higher
notes, and losses are the lowest:

```{r}
husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180)
```

Each game is one note, played in order, 180 notes per minute. Listen for
the long run of high notes during the 20-0 start, then the first low note:
a three-point loss at Michigan in late January.

The pitches snap to a **pentatonic scale**, the five-note scale you get
from only the black keys of a piano. Any data sounds reasonably musical
on it. The real numbers are still there in `notes()`.

### Your choices change what people hear

Just like the axes and colors of a chart, the choices you make shape what a
listener takes away.

Turn `scale` off and pitches follow the numbers exactly. It's more
precise, but harder to listen to:

```{r}
husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180, scale = "none")
```

Narrow the `range` of notes and a dominant season sounds a lot more
ordinary:

```{r}
husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180, range = c("C4", "G4"))
```

You can map loudness, too. Here the postseason gets louder: `game_type` has
three levels (regular season, Big Ten tournament and NCAA tournament), and
each level is louder than the one before:

```{r}
husker_games |>
  sonify_data(point_margin, volume = game_type, instrument = "xylophone", bpm = 180)
```

And instead of one note per row, `time = date` places each game on the
calendar. `length = 20` fits the whole season into 20 seconds. Listen for
the gaps: the break around Christmas, and the waits between tournament
rounds.

```{r}
husker_games |>
  sonify_data(point_margin, time = date, length = 20, instrument = "marimba")
```

## Picking an instrument

`instruments()` lists everything you can use:

```{r}
instruments()
```

The first five are soundeR's built-in sounds, which work everywhere. The
rest are real instruments. You can also filter by family:

```{r}
instruments("chromatic percussion")
```

## Saving your sound

`save_sound()` saves a sonification as a file. The ending of the file name
picks the format:

```{r, eval = FALSE}
season <- husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180)

save_sound(season, "husker-season.mp3") # to share
save_sound(season, "husker-season.wav") # uncompressed audio
save_sound(season, "husker-season.mid") # to open in GarageBand or MuseScore
```

## Things to try

* Which luge podium was closer, men's singles or women's singles? Listen
  first, then check with `filter()` and `arrange()`.
* In the luge, map `pitch = rank` and add `reverse = TRUE` so the winner
  plays the highest note. Does that make it easier to follow?
* Play the Husker season sorted by `point_margin` with `arrange()`. What
  does the sound tell you now that it didn't before, and what did you lose?
* Pick a column from your own data. What should be high, what should be
  loud, and what should decide when a note plays?
