---
title: "Your first sonification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Your first sonification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

A chart turns numbers into shapes you can see. A **sonification** turns
numbers into sounds you can hear. Higher numbers can become higher notes,
the gap between two events can become the silence between two notes, and a
whole season of games can become a short tune.

soundeR does this with one function, `sonify_data()`. It works like the
rest of the tidyverse: give it a data frame, tell it which columns to
listen to, and print the result. In RStudio or Positron an audio player
appears in the Viewer pane. In a Quarto or R Markdown document, like this
one, the player appears on the page.

```{r setup, message = FALSE}
library(soundeR)
library(dplyr)
```

## Real instruments (one-time setup)

soundeR can play 128 real instruments, from piano to xylophone to cello.
They need a one-time download of instrument sounds, about 30 MB:

```{r, eval = FALSE}
sound_setup()
```

If you skip this, everything still works. soundeR uses its own built-in
sounds instead and tells you when it does.

## Fractions of a second

In 2010, during the Vancouver Winter Olympics, The New York Times published
"Fractions of a Second: An Olympic Musical." For each event, it played one
note for every athlete at the moment they crossed the finish line, spaced
by how far they finished behind the winner. You could *hear* how close the
races were.

We can rebuild that with `luge_finals`, the results of every luge event at
the 2026 Winter Olympics in Milan and Cortina d'Ampezzo:

```{r}
luge_finals |>
  select(event, rank, athlete, country, total_time, behind)
```

`behind` is how many seconds each sled finished behind the winner of its
event. Let's listen to the men's singles final:

```{r}
mens_singles <- luge_finals |>
  filter(event == "Men's singles")

mens_singles |>
  sonify_data(time = behind, time_scale = 1, instrument = "piano")
```

There's no pitch here. Every note is the same. Only two things matter:

* `time = behind` decides **when** each note plays. The winner, 0 seconds
  behind, plays first.
* `time_scale = 1` plays it in **real time**: one second of data is one
  second of sound.

Max Langenhan won by 0.596 seconds, and you can hear the gap before the
second note. Now try the men's doubles, where gold and silver were
separated by 0.068 seconds. In real time those first notes nearly blur
together, so slow it down four times with `time_scale = 4`:

```{r}
luge_finals |>
  filter(event == "Men's doubles") |>
  sonify_data(time = behind, time_scale = 4, instrument = "piano")
```

To hear every event, one after another like the original, use
`sequence`. It plays each group in turn, with a second of silence between
them:

```{r}
luge_finals |>
  sonify_data(time = behind, time_scale = 1, sequence = event, instrument = "piano")
```

### Working it out yourself

`luge_finals` already has a `behind` column, but real data usually won't.
You can compute it with `group_by()` and `mutate()`:

```{r}
luge_finals |>
  group_by(event) |>
  mutate(seconds_behind = total_time - min(total_time)) |>
  sonify_data(time = seconds_behind, time_scale = 1, sequence = event, instrument = "piano")
```

Grouping your data doesn't change the sound by itself. `sequence = event`
is what plays the events one at a time. If you pass grouped data without
`sequence`, soundeR reminds you of that.

## What did the data become?

Every row became a note. `notes()` shows you exactly which one, the same way
you'd look at the data behind a chart:

```{r}
race <- mens_singles |>
  sonify_data(time = behind, time_scale = 1, instrument = "piano")

notes(race)
```

`row` matches the row of your data, `onset` is the second the note starts,
`note` is its name (C5 is the C an octave above middle C), and `value` is
the number it came from.

## A season you can hear

`husker_games` holds every game of Nebraska men's basketball's 2025-26
season, the best in program history: 28 wins, a 20-0 start, and the
program's first NCAA tournament wins.

```{r}
husker_games |>
  select(game_number, date, opponent, husker_score, opponent_score, point_margin)
```

This time, let's map `point_margin` to **pitch**. Bigger wins are higher
notes, and losses are the lowest:

```{r}
husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180)
```

Each game is one note, played in order, 180 notes per minute. Listen for
the long run of high notes during the 20-0 start, then the first low note:
a three-point loss at Michigan in late January.

The pitches snap to a **pentatonic scale**, the five-note scale you get
from only the black keys of a piano. Any data sounds reasonably musical
on it. The real numbers are still there in `notes()`.

### Your choices change what people hear

Just like the axes and colors of a chart, the choices you make shape what a
listener takes away.

Turn `scale` off and pitches follow the numbers exactly. It's more
precise, but harder to listen to:

```{r}
husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180, scale = "none")
```

The scale sets the mood, too. `scale = "sad"` plays the same shape
in a minor key, and suddenly a 28-win season sounds like a lament. (The
default is `scale = "happy"`.)

```{r}
husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180, scale = "sad")
```

Narrow the `range` of notes and a dominant season sounds a lot more
ordinary:

```{r}
husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180, range = c("C4", "G4"))
```

You can map loudness, too. Here the postseason gets louder: `game_type` has
three levels (regular season, Big Ten tournament and NCAA tournament), and
each level is louder than the one before:

```{r}
husker_games |>
  sonify_data(point_margin, volume = game_type, instrument = "xylophone", bpm = 180)
```

`duration` sets how long each note rings. A number sets every note to that
many seconds, so `duration = 0.1` makes short, crisp notes that are easier
to tell apart at a fast tempo. A column works too: bigger values ring
longer.

```{r}
husker_games |>
  sonify_data(point_margin, duration = 0.1, instrument = "xylophone", bpm = 180)
```

And instead of one note per row, `time = date` places each game on the
calendar. `length = 20` fits the whole season into 20 seconds. Listen for
the gaps: the break around Christmas, and the waits between tournament
rounds.

```{r}
husker_games |>
  sonify_data(point_margin, time = date, length = 20, instrument = "marimba")
```

## Picking an instrument

`instruments()` lists everything you can use:

```{r}
instruments()
```

The first five are soundeR's built-in sounds, which work everywhere. The
rest are real instruments. `low` and `high` show each one's usual range:
unless you set `range` yourself, notes stay inside it, so a tuba plays low
and a piccolo plays high. You can also filter by family:

```{r}
instruments("chromatic percussion")
```

## More than one voice

So far every note has come from one instrument. There are two ways to use
more.

### Two columns at once

List more than one column in `pitch` and each row plays one note per column
at the same moment. Here Nebraska's score is the marimba and the
opponent's score is the cello:

```{r}
husker_games |>
  sonify_data(c(husker_score, opponent_score), instrument = c("marimba", "cello"), bpm = 150)
```

Both columns share one pitch scale, so 70 points is the same note whichever
team scored it. That's what makes the comparison audible: in a close game,
like the 58-56 win over Michigan State on Jan. 2, the two notes are nearly
the same. In a blowout, like the 90-55 win over Oregon on Jan. 13, they're
far apart.

Put on headphones and you'll also hear the two voices in different places:
with more than one voice, soundeR spreads them from left to right. The
`pan` column in `notes()` says where each note sits, from -1 (left) to 1
(right). Use `pan = "left"`, `pan = "right"` or any number in between to
place notes yourself, `pan = 0` to keep everything in the center, or map a
column, like `pan = location`, to spread groups across the stereo field.

`notes()` shows which voice each note belongs to:

```{r}
husker_games |>
  sonify_data(c(husker_score, opponent_score), instrument = c("marimba", "cello")) |>
  notes() |>
  select(row, voice, onset, note, pan, instrument, value)
```

### An instrument for each group

To give each *kind* of row its own sound, map a column to `voice` and name
an instrument for each group. Here wins are the marimba and losses the
cello:

```{r}
husker_games |>
  sonify_data(point_margin, voice = result,
              instrument = c(W = "marimba", L = "cello"), bpm = 150)
```

`voice` only changes which instrument plays a note, not when it plays. The
games still play in order. (Here the instrument repeats what pitch already
says, since losses are the low notes anyway. Saying the same thing two ways
can make a pattern easier to hear, just as color and position can double
up in a chart.)

If you leave a group out of the list, soundeR plays it on the piano and
tells you which groups it filled in.

## Saving your sound

`save_sound()` saves a sonification as a file. The ending of the file name
picks the format:

```{r, eval = FALSE}
season <- husker_games |>
  sonify_data(point_margin, instrument = "xylophone", bpm = 180)

save_sound(season, "husker-season.mp3") # to share
save_sound(season, "husker-season.wav") # uncompressed audio
save_sound(season, "husker-season.mid") # to open in GarageBand or MuseScore
```

## Making a video

Audio players don't work everywhere. GitHub, most social media sites and
some slide software won't play them. `sonify_video()` turns a sonification
into an MP4 video: a chart of your data, with a playhead that moves across
it and dots that fill in as each note plays.

```{r, eval = FALSE}
luge_finals |>
  sonify_data(time = behind, time_scale = 1, sequence = event, instrument = "piano") |>
  sonify_video(
    "luge.mp4",
    title = "Fractions of a second: luge at the 2026 Winter Olympics",
    subtitle = "Each note is a sled crossing the finish line, in real time",
    x_label = "Seconds behind the winner"
  )
```

With `sequence`, each group gets its own row and the one that's playing is
highlighted. Without it, the chart plots your `pitch` column against time.
With more than one voice, each voice gets its own color.

The chart is a ggplot, so you can change its look with any ggplot2 theme,
the same way you would with a ggplot:

```{r, eval = FALSE}
husker_games |>
  sonify_data(c(husker_score, opponent_score), instrument = c("marimba", "cello")) |>
  sonify_video(
    "husker-scores.mp4",
    title = "Nebraska vs. opponents, 2025-26",
    x_label = "Game",
    y_label = "Points",
    theme = ggplot2::theme_classic(base_size = 16)
  )
```

Titles line up with the left edge of the whole image, not the plot panel,
even with a theme like `theme_classic()`. Making a video takes about as
long as the sound itself: a 30-second sonification takes about 30 seconds
to draw.

To change more than the title, labels and theme, draw the chart yourself
with `sonify_plot()`. It returns an ordinary ggplot, with every note
played, so you can add scales, annotations or anything else. Then hand it
to `sonify_video()` with `plot =`, and the video adds the moving playhead
on top:

```{r, eval = FALSE}
race <- luge_finals |>
  sonify_data(time = behind, time_scale = 1, sequence = event, instrument = "piano")

p <- sonify_plot(race) +
  ggplot2::labs(title = "Fractions of a second", x = NULL) +
  ggplot2::scale_x_continuous(
    breaks = 0:8,
    labels = function(s) ifelse(s == 0, "Winner", paste0("+", s, " s"))
  ) +
  ggplot2::annotate("text", x = 1.9, y = "Men's doubles",
                    label = "0.068 s between gold and silver", hjust = 0, vjust = -1.2)

sonify_video(race, "luge.mp4", plot = p)
```

`sonify_plot()` on its own is handy too, for a picture of your sonification
in a report.

## Things to try

* Which luge podium was closer, men's singles or women's singles? Listen
  first, then check with `filter()` and `arrange()`.
* In the luge, map `pitch = rank` and add `reverse = TRUE` so the winner
  plays the highest note. Does that make it easier to follow?
* Play the Husker season sorted by `point_margin` with `arrange()`. What
  does the sound tell you now that it didn't before, and what did you lose?
* Give each `location` (home, away and neutral) its own instrument with
  `voice`. Can you hear whether Nebraska played differently on the road?
* In the two-score version, try `scale = "chromatic"`. The notes clash
  more. Does that make close games easier or harder to hear?
* Pick a column from your own data. What should be high, what should be
  loud, and what should decide when a note plays?
